---
title: "CausalCoxMGM"
author: "Tyler C. Lovelace"
output:
  github_document:
    keep_md: TRUE
---

# CausalCoxMGM
Implementation of CausalCoxMGM algorithm and scripts for analysis of simulated and real-world biomedical datasets.

## Installation

To install CoxMGM and CausalCoxMGM, run the following command in the terminal:

```
R CMD INSTALL rCausalMGM
```

or alternatively:

```
R CMD INSTALL rCausalMGM/rCausalMGM_1.0.tar.gz
```

## Demonstration of CausalCoxMGM with the WHAS500 dataset

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

First, we begin by loading the necessray R packages for this analysis.

```{r load, echo=TRUE}
library(rCausalMGM)
library(survival)
library(dplyr)
library(survminer)
library(ggplot2)
```

Next, we load the WHAS500 dataset and format the data appropriately for analysis by CausalCoxMGM. First, we create our censored variables, Survival and LengthOfStay, using the `Surv` function from the `survival` R package. Then, we exclude features that are not clinical measurements or patient outcomes, and convert discrete variables into factors.

```{r dataprep, echo=TRUE}
data <- read.csv('cvd/whas500.csv', row.names=1)

data$LengthOfStay <- Surv(data$los, 1-data$dstat)

data$Survival <- Surv(data$lenfol, data$fstat)

data <- data %>% select(-c("admitdate", "disdate", "fdate", "dstat", "lenfol", "fstat", "year", "los"))

data <- data %>% mutate_at(c("cvd", "afb", "sho", "chf", "av3", "gender", "miord", "mitype"), factor)

head(data)
```

Next, we perform stratified 5-fold cross-validation to select model hyperparameters. Cross-validation folds are stratified to have approximately the same number of all-cause mortality and hospital discharge events. We measure model performance based on the total deviance of the censored outcomes. Rather than performing a grid search over the hyperparameters, we perform a random search.

```{r modelselectcv, echo=TRUE, figure=TRUE}
set.seed(43)

idx00 <- which(data$Survival[,2]==0 & data$LengthOfStay[,2]==0)
idx10 <- which(data$Survival[,2]==1 & data$LengthOfStay[,2]==0)
idx01 <- which(data$Survival[,2]==0 & data$LengthOfStay[,2]==1)
idx11 <- which(data$Survival[,2]==1 & data$LengthOfStay[,2]==1)

foldid <- rep(0, 500)
foldid[idx00] <- sample(((1:length(idx00))-1) %% 5 + 1)
foldid[idx10] <- sample(((1:length(idx10))-1) %% 5 + 1)
foldid[idx01] <- sample(((1:length(idx01))-1) %% 5 + 1)
foldid[idx11] <- sample(((1:length(idx11))-1) %% 5 + 1)

table(foldid, data$Survival[,2])
table(foldid, data$LengthOfStay[,2])

lambdas <- runif(100, 0.05, 0.5)

alphas <- runif(100, 0.01, 0.25)

loglik <- matrix(0, 100, 5)
size <- matrix(0, 100, 5)
for (k in 1:5) {
    ig.path <- coxmgmPath(data[foldid!=k,], lambdas=lambdas, rank=F)
    idx <- 0
    for (ig in ig.path$graphs) {
        idx <- idx + 1
        g <- fciStable(data[foldid!=k,], initialGraph=ig,
                       alpha=alphas[idx], orientRule="maxp", rank=F)
        mb <- g$markov.blankets$Survival
        size[idx,k] <- length(mb)
        if (length(mb)==1) {
            mb <- c(1)
        }
        f <- as.formula(paste("Survival ~", paste(mb, collapse=" + ")))
        res <- coxph(f, data[foldid!=k,])
        test.risk <- predict(res, newdata=data[foldid==k,])
        res.test <- coxph(Survival ~ offset(test.risk), data[foldid==k,])
        loglik[idx,k] <- -as.numeric(logLik(res.test))

        mb <- g$markov.blankets$LengthOfStay
        size[idx,k] <- size[idx,k] + length(mb)
        if (length(mb)==1) {
            mb <- c(1)
        }
        f <- as.formula(paste("LengthOfStay ~", paste(mb, collapse=" + ")))
        res <- coxph(f, data[foldid!=k,])
        test.risk <- predict(res, newdata=data[foldid==k,])
        res.test <- coxph(LengthOfStay ~ offset(test.risk), data[foldid==k,])
        loglik[idx,k] <- loglik[idx,k] + -as.numeric(logLik(res.test))
    }
}

sizeMean <- rowMeans(size)
loglikMean <- rowMeans(loglik)
loglikSd <- apply(loglik, 1, sd)

plot(sizeMean, loglikMean, pch=19, col='red')


minIdx <- which.min(loglikMean)
```

Now that we have selected the best set of hyperparameters, we learn the final causal graphical model of all-cause mortality and hospital discharge after hospitilazation with acute myocardial infarction. This is done in two stages: first, we learn the undirected CoxMGM that serves as an initial estimate of the adjacencies in the causal graph. Second, we use FCI-Max to prune adjacencies and orient edges.

```{r finalmodel, echo = TRUE}
ig.path <- coxmgmPath(data, lambda=lambdas)

g <- fciStable(data, initialGraph=ig.path$graphs[[minIdx]],
               alpha=alphas[minIdx], verbose=T, orientRule="maxp")

g
```

Finally, we can generate a simple plot of the causal graphical model in R.

```{r finalmodelplot, echo=TRUE, figure=TRUE}
plot(g, nodeAttr=list(fontsize=36))
```